\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-02-22 europa-basics.tex
updates to europa-basics and concepts

\rx is an adaptive, artificial intelligence based controller and
provides general framework for building reasoning systems for
real-world autonomous vehicles. At MBARI \rx is used for AUV control;
another instantiation of the system is being used for control of a
terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The
development of \rx has been targeted at surveying a number of
oceanographic features which are dynamic and unpredictable
spatio-temporally. Typically this requires our AUVs to balance the
goals of coverage spatially while opportunistically following features
of scientific interest and to do so while being operationally aware of
its own limitations in terms of resources (\eg the battery state of
charge) and overall proximity to other observational assets for
obtaining scientific ground-truth. We are interested in
representational frameworks that allow robots to pursue long-term
objectives while choosing short-term gain and can gracefully deal with
exogenous or endogenous change.

To enable this responsiveness to external observations, the agent has
to be able to synthesize plans insitu and to re-plan. We use the
temporal constraint-based planner \eu with a demonstrated legacy of
having flown on NASA space missions \cite{mus98,jonsson00,bresina05,
  barreiro09}. Our autonomy architecture brings three key innovations
for AUV adaptation: the use of flexible plan representations,
compositional control with the use of partitioned networks and a
demonstrated capability at sea in novel ocean observation methods.

The next few sections will cover \eu in a bottom-up fashion, going
from the generic theoretical frameworks that underlie \eu, to the
specific representation, modeling and algorithms implemented by
it. First, a quick introduction to Constraint Programming and
Constraint-Based Planning is provided.  Next, the specific
representation that \eu uses to perform Constraint-Based Planning is
presented.  After that comes an overview of \eu's architecture and how
it can be embedded in a planning application such as \rx. Finally,
\eu's modeling, inference and search mechanisms are covered in detail
so that the reader can understand how \rx can perform deliberative
planning and react to execution events by embedding and customizing
\eu.

@@

To articulate the fundamentals of automated planning briefly and use
that to motivate the mechanisms we use in our specific form of the
technique we start with a simple example.

@@

At the core of the \rx framework, is the deliberation engine, \eu,
which has a rich legacy from NASA missions. \eu is a versatile
Constraint-based temporal planner which continues to be deployed on a
diverse set of applications. We motivate this section with some
problem domains this planner can handle and then delve in substantial
detail.

\paragraph \textit{Constraint Satisfaction}: A canonical problem in
dealing with constraints is the $N$-Queens problem in which chess queens
must be placed on an  $N$x$N$ chessboard so no queens attack the other. Fig.
\ref{fig:nqueens-1} shows an example of a random positioning of queens
on a  $N$x$N$ chessboard. Queens in violation of the non-attack constraint
are highlighted in red.

\begin{figure} \centering
  \includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
  \caption{\small N-Queens problem. Queens in violation of the
    non-attack constraint are highlighted in red.}
\label{fig:nqueens-1}
\vskip+0.1cm
\end{figure}

If we define $N$x$N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in [1,N]$,
Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a Queen, $0$
otherwise. Then the following constraints need to be satisfied:

@@

\paragraph \textit{Scheduling}: In the Resource Constrained Project
Scheduling Problem (RCPSP) \cite{Bruckera99}, a project consisting of
a set of activities must be scheduled in a way that satisfies minimum
and/or maximum temporal separation constraints. The activity schedule
must also respect fixed limits on the availability of resources
required to perform each activity. In addition to satisfying temporal
and resource constraints, it is common for the user to want to
minimize makespan so that the entire project is finished as early as
possible. Fig. \ref{fig:rcpsp-1} shows an example of a a solution
provided by \eu for an RCPSP instance with 10 activities, 5 resources
and 30 temporal constraints.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
\caption{\small A \eu solution to an RCPSP \cite{Bruckera99} problem.}
\label{fig:rcpsp-1}
\vskip+0.1cm
\end{figure}

\paragraph \textit{Planning}: In the Shopping Agent Problem
\cite{russelnorvig} an agent needs to purchase a set of products (milk,
drill, etc) that are available at specific locations (supermarket,
hardware store, etc), the agent is subject to temporal (must complete
tasks by specific deadlines) and resource (fuel, carrying capacity, etc)
constraints. The agent needs to figure out what actions need to be
performed to find and acquire the required items, as well as when to
perform each of those actions. Fig. \ref{fig:shopping-1} shows a
solution produced by \eu for such a problem instance.

@@

As the examples above show, a Planning problem (actions to achieve a
goal) may embed a Scheduling problem (what resources are necessary to
achieve stated goals) and both Planning and Scheduling may embed a
Constraint Satisfaction problem. 
The relationships between Planning, Scheduling and Constraint
Satisfaction have been examined \cite{smith00} and have lead to use of
constraint reasoning in \eu as its innermost building block.

@@

Constraint Satisfaction Programming (also known as Constraint
Programming or \textsf{CP}) is a discipline that provides a generic framework
for representing, solving and making logical inference on
constraints. A complete treatment of this discipline is given in
\cite{marriott98,apt03} and very concise introductions are provided in
\cite{bartak99,lustig01}.

@@

\item \textbf{Bounds propagation}: To infer upper and lower variable
  bounds. For example, from the constraints $x_1 + x_2 \leq 2$ \&
  $x_i \geq 0$, we can infer $[0,2]$ bounds for $x_1$ and $x_2$

\item \textbf{Domain reduction}: To infer a valid set of values for a variable.
  For example, for  constraints allDifferent($x_1,x_2,x_3$), $x_i \geq
  0$ \& $x_i \leq 4$, if $x_1 = 1$ and $x_2 = 3$ we can infer that the
  valid domain for $x_3$ is reduced  to $\{2,4\}$

@@

A set of constraints and variables that describe a problem domain are
typically represented as a Constraint Network, where the variables are
nodes and constraints are arcs between the variables. Having this
network representation in mind, a pair of variables $x,y$ is said to
be arc consistent if for every value in $x$'s domain, there is a value
in $y$'s domain that is consistent with all the constraints that
connect $x$ and $y$.  A naive approach to ensuring arc consistency
consists of cycling through all the variable pairs and performing
constraint propagation until there are no variable domain changes. The
\texttt{AC-3} algorithm \cite{mackworth77}, a popular implementation, improves
over the naive approach by ignoring constraints whose variables were
not affected since the last iteration. Fig.  \ref{fig:constraintnet} shows a set of constraints on 3 variables
($x$, $y$ and $z$) represented as a constraint network, and their domains before arc consistency is enforced.
After the AC-3 algorithm is executed, the domains for the 3 variables
would be pruned to x=$\{2$\}, y=$\{2$\}, z=$\{2$\}.

@@

Arc Consistency and other, more stringent consistency definitions can
be used to efficiently prove that a \textsf{CP} problem is
unsatisfiable. However, they generally cannot prove that a \textsf{CP}
problem is satisfiable and provide specific variable assignments that
constitute solutions. For that, search algorithms like global
backtracking \cite{hooker05} or local search are used; these use
consistency as an efficient way to prune the search space
\cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
\cite{ghallab04}, but often very efficient in practice using a number
of consistency and search algorithms that are well understood.

@@

The most common planning formulations use a propositional
representation, where the state of the world is represented by a set
of propositions (statements that can be true or false), and operators
change the truth values of these propositions \cite{gen87}. Although
these formulations are powerful and have allowed researchers to
develop numerous contributions in automated planning, there are many
classes of problem domains that are difficult to represent using this
formalism. In particular, it is hard to represent time, resources,
mutual exclusion and concurrency using propositions \cite{ghallab04}.
Constraint Programming is a framework that lends itself very well to
represent and reason about all of those elements \cite{ghallab94},
consequently, there has been a good amount of work building automated
planners that take advantage of CP techniques.  The most common
approach in constraint-based planning has been to encode the entire
planning problem as a CP problem, where action choices and
relationships are represented as variables and/or constraints
\cite{do01,vanbeek99,vossen99,wolfman99}.  The advantage of using a
pure CP formulation is that standard CP solvers can be then used to
solve the planning problem, however, this approach also has a couple
of significant drawbacks:

@@

\item Given that action choices and relationships (rules) are expressed
  through variables, the domain descriptions that result from this
  approach are not intuitive and therefore difficult to understand and
  debug

@@

These drawbacks have been addressed by a small number of planning
frameworks like \texttt{Descartes} \cite{joslin95,joslin96},
\texttt{IxTet} \cite{ghallab87,lemai04}, and \texttt{CAIP} (Constraint-Based
Interval and Attribute Planning)\cite{mus94,frank2003}.  These
frameworks take advantage of a CSP representation to deal with time
and resources, at the same time maintaining an explicit representation
of the elements relevant to planning (action choices, causal
relationships, etc). This combination results in a planning approach
that can deal with the type of constraints found in real-life
problems, while allowing for intuitive problem descriptions and a
natural way to express search algorithms and heuristics. \eu is an
implementation of the \texttt{CAIP} framework. In \texttt{CAIP}, planning problems are
specified in terms of:

@@

\item \textbf{Variables and Constraints}: Variables represent Interval
  attributes and Constraints enable a direct representation of
  temporal, resource and any other the restrictions that a valid plan
  must comply with.

\item \textbf{Domain Configuration Rules}: this is an explicit
  representation of the conceptual relationships between actions and
  states in the problem domain, for instance, for a Shopping Agent to
  be able to purchase a product, it must be at a location where the
  product is available.
\end{enumerate}

>>>>

\rx is an adaptive, artificial intelligence based controller and
provides general framework for building reasoning systems for
real-world autonomous vehicles. At MBARI \rx is used for AUV control;
another instantiation of the system is being used for control of a
terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The
development of \rx has been targeted at surveying a number of
oceanographic features which are dynamic and spatio-temporally
unpredictable. Typically this requires our AUVs to balance the goals
of coverage spatially while opportunistically following features of
scientific interest and to do so while being aware of its own
limitations in terms of resources (\eg the battery state of charge)
and overall proximity to other observational assets for obtaining
scientific ground-truth. We are interested in representational
frameworks that allow robots to pursue long-term objectives while
choosing short-term gain and can gracefully deal with exogenous or
endogenous change.

To enable this responsiveness to external observations, the agent has
to be able to synthesize plans insitu and to re-plan without human
intervention. To do so we use a temporal constraint-based planner with
a demonstrated legacy of having flown on NASA space missions. Our
autonomy architecture brings three key innovations for AUV control:
in-situ plan synthesis, the use of flexible plan representations with
a sound theoretical basis and systematic compositional control with
the use of partitioned networks doing so with a demonstrated
capability at sea in novel ocean observation methods.

The next few sections will cover \eu in a bottom-up fashion, going
from the generic theoretical framework that underlies \eu, to the
specific representation, modeling and algorithms implemented by it.  A
brief introduction to Constraint Programming and Constraint-Based
Planning is provided followed by the specific representation that \eu
uses to undertake Constraint-Based Planning.  Subsequently, we provide
an overview of \eus architecture and how it can be embedded in a
planning application such as \rx. Finally, \eus modeling, inference
and search mechanisms are covered in detail so that the reader can
understand how \rx can perform deliberative planning and react to
execution events by embedding and customizing \eu.

@@

To articulate the fundamentals of automated planning briefly and use
that to motivate the mechanisms we use in our specific form of
planning we start with a simple example.

@@

At the core of the \rx framework, is the deliberation engine, \eu,
which has a rich legacy from NASA missions \cite{mus98,rajan00,
  jonsson00,aichang04, bresina05}. \eu is a versatile Constraint-based
Temporal Planner which continues to be deployed on a diverse set of
applications including recently, planning for the International Space
Station \cite{barreiro09}. We motivate this section with some problem
domains this planner can handle and then proceed to go into detail.

\paragraph \textit{Constraint Satisfaction}: A canonical problem in
dealing with constraints is the $N$-Queens problem in which chess queens
must be placed on an  $N$x$N$ chessboard so no queens attack the other. % Fig.
% \ref{fig:nqueens-1} shows an example of a random positioning of queens
% on a  $N$x$N$ chessboard. Queens in violation of the non-attack constraint
% are highlighted in red.
% \begin{figure} \centering
%   \includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
%   \caption{\small N-Queens problem. Queens in violation of the
%     non-attack constraint are highlighted in red.}
% \label{fig:nqueens-1}
% \vskip+0.1cm
% \end{figure}
If we define $N$x$N$ variables Q$_{rc}$, $r \in [1,N]$, $c \in [1,N]$,
Q$_{rc} = 1$ if cell $r,c$ in the chessboard is occupied by a Queen,
$0$ otherwise. Then the following constraints need to be satisfied:

@@

\paragraph \textit{Scheduling}: In the Resource Constrained Project
Scheduling Problem (\texttt{RCPSP}) \cite{Bruckera99}, a project
consisting of a set of activities must be scheduled in a way that
satisfies minimum and/or maximum temporal separation constraints. The
activity schedule must also respect fixed limits on the availability
of resources required to perform each activity. In addition to
satisfying temporal and resource constraints, it is common for the
user to want to minimize makespan \cite{ghallab04} so that the entire
project is finished as early as possible. Fig. \ref{fig:rcpsp-1} shows
an example of a a solution provided by \eu for an \texttt{RCPSP}
instance with 10 activities, 5 resources and 30 temporal constraints.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
\caption{\small A \eu solution to an \texttt{RCPSP} \cite{Bruckera99}
  problem.}
\label{fig:rcpsp-1}
\vskip+0.1cm
\end{figure}

\paragraph \textit{Planning}: In the Shopping Agent Problem
\cite{russelnorvig} an agent needs to purchase a set of products
(milk, drill, etc) that are available at specific locations
(supermarket, hardware store, etc) and the agent is subject to
temporal (must complete tasks by specific deadlines) and resource
(fuel, carrying capacity, etc) constraints. The agent needs to figure
out what actions need to be performed to find and acquire the required
items, as well as when to perform each of those
actions. Fig. \ref{fig:shopping-1} shows a solution produced by \eu
for such a problem instance.

@@

As the examples above show, a Planning problem (actions to achieve a
goal) may embed a Scheduling problem (what resources are necessary to
achieve stated goals) and both Planning and Scheduling may embed a
Constraint Satisfaction problem.  The relationships between Planning,
Scheduling and Constraint Satisfaction have been examined
\cite{smith00} and have lead to use of constraint reasoning in \eu as
a core building block.

@@

Constraint Satisfaction Programming (also known as Constraint
Programming or \textsf{CP}) is a discipline that provides a generic
framework for representing, solving and making logical inference on
constraints. A complete treatment of this discipline is given in
\cite{marriott98,apt03} with concise introductions provided in
\cite{bartak99,lustig01}.

@@

\item \textbf{Bounds propagation} infers upper and lower variable
  bounds. For example, from the constraints $x_1 + x_2 \leq 2$ \& $x_i
  \geq 0$, we can infer $[0,2]$ bounds for $x_1$ and $x_2$.

\item \textbf{Domain reduction} infers a valid set of values for a
  variable.  For example, for constraints allDifferent($x_1,x_2,x_3$),
  $x_i \geq 0$ \& $x_i \leq 4$, if $x_1 = 1$ and $x_2 = 3$ we can
  infer that the valid and reduced domain for $x_3$ is $\{2,4\}$.

@@

A set of constraints and variables that describe a problem domain are
typically represented as a \emph{Constraint Network}, where the
variables are nodes and constraints are arcs between the variables. A
pair of variables $x \& y$ is said to be \emph{arc consistent} if for
every value in $x$'s domain, there is a value in $y$'s domain that is
consistent with all the constraints that connect $x$ and $y$.  A naive
approach to ensuring arc consistency consists of cycling through all
the variable pairs and performing constraint propagation until there
are no variable domain changes. The \texttt{AC-3} algorithm
\cite{mackworth77}, a popular implementation, improves over the naive
approach by ignoring constraints whose variables were not affected
since the last iteration. Fig. \ref{fig:constraintnet} shows a set of
constraints on $3$ variables ($x$, $y$ and $z$) represented as a
constraint network, and their domains before arc consistency is
enforced.  After the \texttt{AC-3} algorithm is executed, the domains
for the variables would be pruned to x=$\{2$\}, y=$\{2$\}, z=$\{2$\}.

@@

Arc Consistency and other more stringent consistency definitions can
be used to efficiently prove that a \textsf{CP} problem is
unsatisfiable. However, they generally cannot prove that a \textsf{CP}
problem is satisfiable and provide specific variable assignments that
constitute solutions. To do so, search algorithms like global
backtracking \cite{hooker05} or local search are used; these use
consistency as an efficient way to prune the search space
\cite{cp06}. In theory, solving a \textsf{CP} problem is NP-Hard
\cite{ghallab04}, but often in practice a number of consistency and
search algorithms can be used.

@@

The most common planning formulations use a propositional
representation, where the state of the world is represented by a set
of propositions (statements that can be true or false), and operators
change the truth values of these propositions \cite{gen87}. Although
these formulations are powerful and have allowed researchers to
develop numerous contributions in automated planning, there are many
classes of problem domains that are difficult to represent using this
formalism. In particular, it is hard to represent time, resources,
mutual exclusion and concurrency using propositions \cite{ghallab04}.
Constraint Programming is a framework that lends itself well to
represent and reason about all of those elements \cite{ghallab94}.
Consequently, there has been substantial effort in building automated
planners that take advantage of \textsf{CP} techniques.  The most
common approach in constraint-based planning has been to encode the
entire planning problem as a \textsf{CP} problem, where action choices
and relationships are represented as variables and/or constraints
\cite{do01,vanbeek99,vossen99,wolfman99}.  The advantage of using a
pure \textsf{CP} formulation is that standard solvers can be then used
to solve the planning problem. However, this approach also has a
couple of significant drawbacks:

@@

\item Given that action choices and relationships (rules) are
  expressed through variables, the domain descriptions that result
  from this approach are not intuitive and therefore difficult to
  understand and debug.

@@

These drawbacks have been addressed by a number of planning frameworks
like \texttt{Descartes} \cite{joslin95,joslin96}, \texttt{IxTet}
\cite{ghallab87,lemai04} and \texttt{CAIP} (Constraint-Based Interval
and Attribute Planning) \cite{mus94,frank2003}.  These frameworks take
advantage of a CSP representation to deal with time and resources, at
the same time maintaining an explicit representation of the elements
relevant to planning (action choices, causal relationships, etc). This
combination results in a planning approach that can deal with the type
of constraints found in real-world problems while allowing for
intuitive problem descriptions and a natural way to express search
algorithms and heuristics. \eu is one such implementation of the
\texttt{CAIP} framework, in which planning problems are specified in
terms of:

@@

\item \textbf{Variables and Constraints}: Variables represent Interval
  attributes and constraints enable a direct representation of
  temporal, resource and any other restrictions that a valid plan must
  comply with.

\item \textbf{Domain Configuration Rules}: This is an explicit
  representation of the conceptual relationships between actions and
  states in the problem domain, for instance, for a Shopping Agent to
  be able to purchase a product, it must be at a location where the
  product is available.

### 2012-02-22 intro.tex
updates to europa-basics and concepts

This chapter is organized as follows. We start with some foundational
concepts in our Planning framework in Section~\ref{sec:concepts} which
includes a brief overview of AI Planning in
Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
deeper into the \eu framework itself, the basis of all deliberation
within \rx. A key portion is Section~\ref{sec:europa:pr} which details
the Internal plan representation essential to planning within a
situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
Inference (Section~\ref{sec:europa:inference}) and Search
(Section~\ref{sec:europa:search}) further detail the important details
associated with \eu for the reader. With these foundational elements
out of the way, we transition to \rx in Section~\ref{sec:arch}, bring
out architectural details in Section~\ref{sec:arch:trex}, deal with
the execution cycle in Section~\ref{sec:arch:exec}, highlight how \rx
deliberates in Section~\ref{sec:arch:europa} finally delving deeper
into the key concept of synchronization in
Section~\ref{sec:arch:synch}. Further details on deliberation within
\rx (Section~\ref{sec:arch:plan}) and how planning and execution are
interleaved are shown in Section~\ref{sec:arch:intertwine}. We offer
results from our experiments at sea in Section~\ref{sec:results},
offer some insights into the near-term future goals of our research in
Section~\ref{sec:future} and conclude with
Section~\ref{sec:conclusion}.

>>>>

This chapter is organized as follows. We start with some foundational
concepts in our Planning framework in Section~\ref{sec:concepts} which
includes a brief overview of AI Planning in
Section~\ref{sec:planningfound}. In Section~\ref{sec:basics} we go
deeper into the \eu framework itself, the basis of all deliberation
within \rx. A key portion is Section~\ref{sec:europa:pr} which details
the internal plan representation essential to planning within a
situated agent like \rx. Modeling (Section~\ref{sec:europa:modeling}),
Inference (Section~\ref{sec:europa:inference}) and Search
(Section~\ref{sec:europa:search}) further detail the important details
associated with \eu for the and relevant to how \rx works to the
reader. With these foundational elements out of the way, we transition
to \rx itself, in Section~\ref{sec:arch}, bring out architectural
details in Section~\ref{sec:arch:trex}, deal with the execution cycle
in Section~\ref{sec:arch:exec}, highlight how \rx deliberates in
Section~\ref{sec:arch:europa} finally delving deeper into the key
concept of synchronization in Section~\ref{sec:arch:synch}. Further
details on deliberation within \rx (Section~\ref{sec:arch:plan}) and
how planning and execution are interleaved are shown in
Section~\ref{sec:arch:intertwine}. We offer results from our
experiments at sea in Section~\ref{sec:results}, offer some insights
into the near-term future goals of our research in
Section~\ref{sec:future} and conclude with
Section~\ref{sec:conclusion}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "setobook"
%%% End:
